---
title: "Motor Trend MPG Analysis"
author: "Altamash Ali"
date: "20/01/2026"
output: pdf_document
---

## Executive Summary

This report examines the relationship between transmission type (automatic vs. manual) and fuel efficiency (MPG) using the `mtcars` dataset. Our analysis shows that manual transmissions are associated with higher MPG. While a simple comparison shows a 7.2 MPG advantage for manual, after adjusting for weight and horsepower, the manual advantage is approximately 2.94 MPG.

## Exploratory Data Analysis

We look at the distribution of MPG by transmission type (0 = automatic, 1 = manual).

```{r echo=FALSE}
data(mtcars)
mtcars$am <- factor(mtcars$am, labels = c("Auto", "Manual"))
boxplot(mpg ~ am, data = mtcars, col = c("red", "blue"), ylab = "MPG", main = "MPG by Transmission")
```

### Model Selection and Results

We compare a simple model to a multivariable model that includes weight (wt) and horsepower (hp) to avoid bias.

```{r}
fit_simple <- lm(mpg ~ am, data = mtcars)
fit_multi <- lm(mpg ~ am + wt + hp, data = mtcars)
summary(fit_multi)$coef
```

Interpretation: Holding weight and horsepower constant, manual transmission increases fuel efficiency by 2.94 MPG. The p-value for transmission is significant, and the R-squared shows the model explains 84% of the variance.

### Conclusion

Manual transmission is better for MPG than automatic transmission. The quantified difference is an increase of 2.94 MPG when accounting for the fact that manual cars are generally lighter and have different horsepower profiles.

### Appendix: Residual Diagnostics

```{r}
par(mfrow = c(2, 2))
plot(fit_multi)
```

---

### Source Code

The R Markdown source code used to generate this report can be found at the following GitHub repository:  
https://github.com/realAltamashAli/Motor_Trend_MPG_Analysis
